Bayesian phylogenetic estimation of fossil ages reveals remarkable consistency of rocks and clocks

### 2015-08-26 fossilDating.tex
adding canid analysis

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
\bibliography{fossilDating.bib}

@@

@@

Contention between palaeontologists and molecular biologists over which data provides the most accurate inferences about evolutionary history has previously fostered an adversarial relationship between the two fields \autocite{Donoghue2007}. 
Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \autocite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \autocite{BentonAyala2003}. 
%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}. 
% AJD: The following two sentences are inflammatory to palaeontologists
%Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. 
%Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang:2006yu}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}. Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \autocite{Donoghue2007}.

@@

We extend this work into the realm of morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters. 
This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 
We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
The method is applied to a rich and well-characterized morphological data set of extant penguins and their fossil ancestors \autocite{ksepka2010,ksepka2012}.

@@

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree or chrono-phylogeny. Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree. We refer to this estimate of a fossil's age as its {\em phylogenetic age}. In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age. These three questions are investigated in the context of a morphological data set of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}.

@@

For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,160)$ Mya. We chose 160 as the upper limit as we expect no serious argument that any bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny. For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}. We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.

@@

We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil. In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,160)$ Mya. The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,160)$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1, and the turnover $r$ and sampling probabiltiy $s$ have uniform priors on $[0,1]$)

The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
It remains to calculate the probabilities $p(H_1|M)$ and $p(H_2|M)$.
In order to do so, we derive the probability density of sampling a fossil at time $t$ in the past, given the model $M$. 
For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
$$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
We simplify,
\begin{eqnarray*}
p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
\end{eqnarray*}
Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,160)$ on a grid.
Figuring out now how things change in case we assume uniform prior on $N$, and then evaluate $T=\log(N)/d$.

@@

\subsection*{\Mstrict{}: Strict clock model, no rate variation among sites}
Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.9$. The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

@@

\subsection*{\Mrelaxed{}: Relaxed clock, site partitions, rate variation among sites}

@@

\subsection*{Comparison of M1 and M8 results}

Overall the results from the M1 and M8 models were strikingly concordant. Figure \ref{fig:compareM1M8} shows three regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (c) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.

@@

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model. 
There are a diverse array of potential applications for this methodology. 
The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found or a complete lack of provenance data (e.g. a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}). 
It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 
The median error in age estimates for the penguin data set we analysed here was 2 My, using either a very simple or a more complex model of discrete morphological change. In both cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.

>>>>

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
\bibliography{fossilDating.bib}

@@

\newcommand{\todo}[1]{\textcolor{blue}{{TODO: #1}}}

@@

Contention between palaeontologists and molecular biologists over which data provides the most accurate inferences about evolutionary history has previously fostered an adversarial relationship between the two fields \autocite{Donoghue2007}. 
Although there has indeed been much controversy surrounding apparent discrepancies between palaeontological and molecular phylogenetic inferences \autocite{BromhamPenny2003} it is also clear that fossil and molecular data both produce broadly concordant views of evolutionary history \autocite{BentonAyala2003}. 
%Indeed one of the first examples of applying a macroevolutionary phylogenetic model to perform inference directly on fossil occurrence data in fact demonstrated concordance between inferences from  molecular data and paleontological data in primates \autocite{tavare2002using}. 
% AJD: The following two sentences are inflammatory to palaeontologists
%Nevertheless there have been only few attempts to apply phylogenetic reasoning to palaeontological questions. 
%Models of molecular evolution have been refined over many decades \autocite{Felsenstein2004,Yang:2006yu}. 
The continual improvement of models and methods for statistical phylogenetic inference from molecular sequence data is well documented \autocite[e.g.][]{Felsenstein2004,Yang2014}, and in recent years it is arguably the case that molecular phylogenetics has taken primacy over the fossil record in providing a timescale for evolutionary history \autocite{Donoghue2007}. Nevertheless molecular phylogenetic inference of evolutionary timescales relies critically on calibration by the fossil record \autocite{Donoghue2007}.

@@

We extend this work into the realm of morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters. 
This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 
We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.

@@

A previous study described how a set of fossils with discrete morphological characters could be used to estimate a time-tree or chrono-phylogeny. Here we additionally allow for one or more of the fossils to have broad uninformative priors on their age. This allows for the age of some of the fossils to be estimated solely based on their morphological characters and the phylogenetic affinities of their morphology to other fossils with known ages in the time-tree. We refer to this estimate of a fossil's age as its {\em phylogenetic age}. In estimating each of the fossils' phylogenetic ages in turn, three questions can be answered: (i) How much information about an individual fossil's age is available from phylogenetic analysis of morphological data alone, (ii) What is the level of phylogenetic evidence in support of the palaeontological age range for a fossil, and (iii) How does the amount of morphological data available for a new fossil and the number of related reference fossils of known age affect the accuracy of the phylogenetic estimate of a fossil's age. These three questions are investigated using two morphological data sets, one of 36 fossil penguins and their extant relatives \autocite{ksepka2010,ksepka2012,gavryushkina2015bayesian}, and one of canids \autocite{Slater2015}.

@@

For each of the 36 penguin fossils in turn we performed a separate Bayesian phylogenetic analysis in which the focal fossil's palaeontological age constraints were replaced by a broad uniform prior of $(0,T_\text{max})$ Mya. For the penguins we chose $T_\text{max}=160$ as the upper limit as we expect no serious argument that any crown bird fossil would be older than 160Mya and it also corresponds to the upper limit on the origin of the whole phylogeny. For the choices of morphological substitution model and tree prior we followed \textcite{gavryushkina2015bayesian} and considered two of her modelling combinations. The first was \Mstrict{} in which a strict clock and the simplest model of morphological substitution was assumed \autocite{Lewis2001}. We also employed \Mrelaxed{} from \textcite{gavryushkina2015bayesian} which involved a more complex model assuming a relaxed clock model \autocite{Drummond2006}, partitioning of the sites into groups of equal state count and an additional parameter for gamma-distributed rate variation across sites \autocite{yang:1994ma}.

@@

We are interested in computing the Bayes factor that quantifies the amount of phylogenetic evidence in support of the geological age range for each fossil. In this case $H_1$ is the hypothesis that the true fossil age is within the given paleontological age range, and $H_2$ is the alternative hypothesis that the true fossil age is outside the geological range, but within the broader interval of $(0,T_\text{max})$ Mya. The model $M$ is the fossilized birth-death process with the specified priors on its parameters (in our analyses the time of origin $T$ of the process has a uniform prior on $(0,T_\text{max})$, the net diversification rate $d$ has a lognormal prior with mean -3.5 and standard deviation 1.5, and the turnover $r$ and sampling probability $s$ have uniform priors on $[0,1]$)

The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
It remains to calculate the probabilities $p(H_1|M)$ and $p(H_2|M)$.
In order to do so, we derive the probability density of sampling a fossil at time $t$ in the past, given the model $M$. 
For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
$$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
We simplify,
\begin{eqnarray*}
p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
\end{eqnarray*}
Now, in order to visualize the results, we sample $T,d,r,s$ from their priors and evaluate for  $t \in (0,T_\text{max})$ on a grid.
Figuring out now how things change in case we assume uniform prior on $N$, and then evaluate $T=\log(N)/d$.

\subsection*{Applying a morphological clock to Canid evolution}

The second data set that we investigated was a morphological data matrix of 132 canid species \autocite[9 extant and 123 fossil;][]{Slater2015}. 
The 9 extant species represent about 25\% of the extant canid species and include representatives of four genera (6 {\em Canis}, 1 {\em Cuon}, 1 {\em Lycaon}, 1 {\em Urocyon}) and both tribes (8 Canini, 1 Vulpini). 
For this data set we did not have a geological age range for each fossil, but instead the author of the study provided a point estimate for each fossil species' geological age.

We applied the M1 model to this data set assuming a fixed geological age for each taxon. Unlike the original study \autocite{Slater2015} we did not apply any other constraints or priors on ancestral divergence times beyond the ages of the included fossils. 
As with the penguin data set we also performed an analysis for each of the 123 canid fossils in turn in which we relaxed the focal fossil's palaeontological age constraints to a broad uniform prior of $(0,T_\text{max})$ Mya.

@@

\subsection*{Penguins conform well to a strict morphological clock}
Although \Mstrict{} is a very simple model, the phylogenetic age estimates for the penguin fossils were remarkably consistent with their palaeontological age ranges. 
Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic age estimates. The points in this plot have $R^2 = 0.9$. 
The median error (difference between the phylogenetic median and the geological median) is 2.0 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

@@

\subsection*{Relaxing the clock, site partitions, rate variation among sites}

@@

\subsection*{Comparison of simple and complex model results}

Overall the results of analyzing the penguin data set with the M1 and M8 models were strikingly concordant. Figure \ref{fig:compareM1M8} shows four regressions between the two models: (a) Regression of estimated phylogenetic age of \Mstrict{} against \Mrelaxed{}, (b) Regression of the error in the phylogenetic age estimate of \Mstrict{} against \Mrelaxed{} (c) Regression of posterior probability of palaeontological range of \Mstrict{} against \Mrelaxed{}, (d) Regression of Bayes factor (BF) for palaeontological range of \Mstrict{} against \Mrelaxed{}.

\subsection*{Canids conform well to a morphological clock}

The canid data set shows remarkably good consistency between palaeontological ages and phylogenetic ages, even with the simple strict morphological clock model (\Mstrict{}).
The $R^2 = 0.901$ between the phylogenetic and palaeontological ages (see Figure \ref{fig:canidMorphVsGeo}). 
Only 16 out of 123 fossils (13\%) don’t have the paleontological age in the credible interval of the phylogenetic age and there are no extreme outliers. 
The median error is $1.66$My, which is even more accurate than the penguin data set.
There are less characters than the penguins (122 versus $\sim 250$) so the uncertainty in the individual age estimates is greater (median 99\% HPD range = 6.5 My), but overall it seems to work very nicely.

@@

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate and precise age of the fossil based on a phylogenetic model. 
There are a diverse array of potential applications for this methodology. 
The most obvious is the estimation of dates for fossils that are temporally unconstrained, either due to poor knowledge of the age of the sediments in which it was found or a complete lack of provenance data (e.g. a recent fossil described as a `four-legged snake' has excited controversy for a lack of provenance\footnote{See \url{http://news.sciencemag.org/paleontology/2015/07/four-legged-snake-fossil-stuns-scientists-and-ignites-controversy}}). 
It can also be used a way of testing the ``morphological clock'' and to discover potential problems in the data by identifying outlier fossils with respect to model fit. 
The median error in age estimates for the two data sets investigated were 2 My and 1.6 My respectively, using either a very simple or more complex models of discrete morphological change. In all cases we used the new fossilized birth-death tree prior, which is a crucial ingredient in allowing for the estimating of fossil ages under a birth-death tree prior.

\todo{How does the number of fossils in the reference affect the precision and accuracy of the fossil age estimates ?}
